A SAT-Based Counterexample Guided Method for Unbounded Synthesis

### 2016-01-28 paper.tex

At every round of the game, the \emph{environment} picks an
uncontrollable label, the \emph{controller} responds by choosing a controllable 
label and the game transitions into a new state according to $\delta$. 
A \emph{run} of a game $(x_0, u_0, c_0), (x_1, u_1, c_1) \dots (x_n, u_n,
c_n)$ is a chain of state and label pairs s.t.\,$\delta(x_k, u_k, c_k, x_{k+1})$. 
A run is winning for the controller if $x_0 = I \land \forall i \in \{1..n\} (\lnot E(x_i))$. 
In a \emph{bounded game} of rank $n$ all runs are restricted to length $n$, whereas unbounded 
games consider runs of infinite length. Since we consider only deterministic games, a run 
is uniquely described by a list of assignments to $L_u$ and $L_c$.

@@

\subsection{Realisability of Bounded Safety Games}

@@

\subsection{Abstract Game Trees}

\tikzset{every node/.style={solid}}
\tikzstyle{fixed}=[solid]
\tikzstyle{unfixed}=[dash pattern = on 2pt off 2pt]
\begin{figure}
    \centering
    \begin{subfigure}[t]{.25\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] {}
                    edge from parent [fixed] node [right] {$d$}
                }
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \captionsetup{width=.8\textwidth}
        \caption{AGT}
        \label{fig:agt}
    \end{subfigure}%
    \begin{subfigure}[t]{.25\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        node [left=4pt] {$\epsilon$}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    node [left=4pt] {$\beta$}
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] {}
                    node [left=4pt] {$\gamma$}
                    edge from parent [fixed] node [right] {$d$}
                }
                node [right=4pt] {$\alpha$}
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \captionsetup{width=.8\textwidth}
        \caption{AGT with strategy}
        \label{fig:strategy}
    \end{subfigure}%
    \begin{subfigure}[t]{.25\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        node [left=4pt] {$\epsilon$}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    node [left=4pt] {$\beta$}
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] (sroot) {}
                    child {node [below left=0 and -10pt] (sa) {}
                        edge from parent [draw=none]
                    }
                    child {node [below right=0 and -10pt] (sb) {}
                        edge from parent [draw=none]
                    }
                    node [left=4pt] {$\gamma$}
                    edge from parent [fixed] node [right] {$d$}
                }
                node [right=4pt] {$\alpha$}
                node [above=4pt] {$\langle s, k \rangle$};
                \draw[unfixed] (sa.center) -- (sb.center);
                \draw[unfixed] (sroot.south) -- (sa.center);
                \draw[unfixed] (sroot.south) -- (sb.center);
        \end{tikzpicture}
        \captionsetup{width=.8\textwidth}
        \caption{AGT with spoiling strategy}
        \label{fig:spoiling}
    \end{subfigure}%
    \begin{subfigure}[t]{.25\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [right] {$e$}
                    }
                    edge from parent [fixed] node [right] {$d$}
                }
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \captionsetup{width=.8\textwidth}
        \caption{Refined AGT}
        \label{fig:refined}
    \end{subfigure}
    \caption{Abstract game trees}
    \label{fig:alltrees}
\end{figure}

An abstraction of the game restricts actions available to one of the players.
Specifically, we consider abstractions represented as trees of actions,
referred to as \emph{abstract game trees}.  Together with a state-round pair
$\langle s,k\rangle$, an abstract game tree defines an \emph{abstract game}
played from this state.  In the abstract game, the environment player is
required to pick actions from the tree, starting from the root node.  After
reaching a leaf, it continues playing unrestricted.  The tree in
Figure~\ref{fig:agt} restricts the initial environment action to the set $\{a,
d\}$.  After choosing action $d$, the environment reaches a leaf of the tree
and continues playing unrestricted.  Alternatively, after choosing $a$, the
environment is required to play action $b$ in the next round.

Nodes of an abstract game tree are uniquely identified by the list of edge
labels along the path from the root to the node.  We identify an abstract game
tree with the set of its nodes.  For example, the tree in Figure~\ref{fig:agt}
can be written as $\{(),(d),(a),(a,b)\}$.  We denote $\textsc{leaves}(T)$ the
subset of leaf nodes of a tree $T$.

A \emph{partial strategy} $Strat: T \rightarrow L_c $ assigns a controllable
action to be played in each node of the abstract game tree.
Figure~\ref{fig:strategy} shows an example partial strategy.  The controller
starts by choosing action $\alpha$.  If the environment plays $a$, the
controller responds with $\beta$ in the next round, and so on. Given a partial
strategy $Strat$, we can map each leaf $l$ of the abstract game tree to
$\langle s',i'\rangle=\textsc{outcome}(\langle s, i\rangle, Strat, l)$ obtained
by playing all controllable and uncontrollable actions on the path from the
root to the leaf.

@@

The algorithm takes a concrete game $G$ as an implicit argument. In addition it
takes a player (controller or environment), state-round pair $\langle s,
k\rangle$ and an abstract game tree $T$ and returns a winning partial strategy
for $T$, if one exists.  The environment takes the first move in each step of
the game so the algorithm initially plays on the environment's behalf. The
initial invocation takes the initial state $\langle I, n\rangle$ and an empty
abstract game tree $\emptyset$. The empty game tree does not constrain opponent
moves, hence solving such an abstraction is equivalent to solving the original
concrete game.

The algorithm is organised as a counterexample-guided abstraction refinement
(CEGAR) loop.  The first step of the algorithm uses the \textsc{findCandidate}
function, described in detail below, to come up with a candidate partial
strategy for $T$. If it fails to find a strategy, this means that no winning
partial strategy exists for $T$.  If, on the other hand, a candidate partial
strategy is found, we need to verify if it is indeed winning for $T$.

The \textsc{verify} procedure searches for a \emph{spoiling} counterexample
strategy in each leaf of the candidate partial strategy by calling
\textsc{solveAbstract} for the opponent. The dual solver solves games on behalf
of the controller player.  {Figure~\ref{fig:spoiling}} shows a spoiling
strategy discovered one of the leaves of the partial strategy.

@@

The way game trees are refined is determined by the formula passed to the SAT
solver in $\textsc{findCandidate}$. This formula contains CNF encodings of all
of the unrolled runs represented by an abstract game tree and the winning
condition of the current player. Runs are encoded by copying the transition
relation for every step in the abstract game. When playing for the controller,
the SAT solver searches for a satisfying assignment to the unfixed label
variables in tree so that none of the runs reaches the error state. The
environment formulation is satisfiable if any run does reach the error state.
The formula is constructed recursively from the root of a tree by
$\textsc{treeFormula}$.

@@

%%%The bounded synthesis algorithm (Algorithm \ref{alg:bounded}) begins with an
%%%empty game as seen in Figure \ref{fig:emptytree}.  Initially we are playing on
%%%behalf of the environment because it chooses the first move in each step.  The
%%%empty game is passed to the SAT solver, which searches for a candidate
%%%environment strategy.  If a candidate is found (Figure \ref{fig:candidate})
%%%then it is checked for a spoiling strategy by solving for the controller.  If
%%%no spoiling strategy exists, that means our candidate is a winning strategy and
%%%the algorithm terminates.  Otherwise we find a counterexample (Figure
%%%\ref{fig:cex}), which is used to refine the empty game tree to include the
%%%first move from the controller's spoiling strategy (Figure \ref{fig:refined}).
%%%The algorithm continues by finding a new candidate for the environment (Figure
%%%\ref{fig:candidate2}), and a new counterexample to refine the game abstraction
%%%once again (Figure \ref{fig:refined2}). The full procedure is listed in
%%%Algorithm \ref{alg:bounded}.

>>>>

At every round of the game, the \emph{environment} picks an
uncontrollable label, the \emph{controller} responds by choosing a controllable 
label and the game transitions into a new state according to $\delta$. 
A \emph{run} of a game $(x_0, u_0, c_0), (x_1, u_1, c_1) \dots (x_n, u_n,
c_n)$ is a chain of state and label pairs s.t.\,$\delta(x_k, u_k, c_k, x_{k+1})$. 
A run is winning for the controller if $x_0 = I \land \forall i \in \{1..n\} (\lnot E(x_i))$. 
In a \emph{bounded game} with bound $n$ all runs are restricted to length $n$, whereas unbounded 
games consider runs of infinite length. Since we consider only deterministic games, a run 
is uniquely described by a list of assignments to $L_u$ and $L_c$.

@@

@@

\tikzset{every node/.style={solid}}
\tikzstyle{fixed}=[solid]
\tikzstyle{unfixed}=[dash pattern = on 2pt off 2pt]
\begin{figure}
    \centering
    \begin{subfigure}[t]{.25\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] {}
                    edge from parent [fixed] node [right] {$d$}
                }
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \captionsetup{width=.8\textwidth}
        \caption{AGT}
        \label{fig:agt}
    \end{subfigure}%
    \begin{subfigure}[t]{.25\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        node [left=4pt] {$\epsilon$}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    node [left=4pt] {$\beta$}
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] {}
                    node [left=4pt] {$\gamma$}
                    edge from parent [fixed] node [right] {$d$}
                }
                node [right=4pt] {$\alpha$}
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \captionsetup{width=.8\textwidth}
        \caption{AGT with partial strategy}
        \label{fig:strategy}
    \end{subfigure}%
    \begin{subfigure}[t]{.25\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [right] {$e$}
                    }
                    edge from parent [fixed] node [right] {$d$}
                }
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \captionsetup{width=.8\textwidth}
        \caption{Refined AGT}
        \label{fig:refined}
    \end{subfigure}
    \caption{Abstract game trees}
    \label{fig:alltrees}
\end{figure}

An abstraction of the game restricts actions available to one of the players.
Specifically, we consider abstractions represented as trees of actions,
referred to as \emph{abstract game trees}.  Figure~\ref{fig:agt} shows an example abstract game
tree for the controller (abstract game trees for the environment are similar).  In the abstract game, 
the controller is required to pick actions
from the tree, starting from the root node.  After reaching a leaf, it
continues playing unrestricted.  The tree in Figure~\ref{fig:agt} restricts the
initial controller action to the set $\{a, d\}$.  After choosing action $a$,
the controller is required to play action $b$ in the next round.  It then reaches a leaf 
of the tree and continues playing unrestricted.
The root of the tree is annotated by the initial state $s$ of the abstract game and the 
bound $k$ on the number of rounds. 
We denote $\textsc{nodes}(T)$ the set of all nodes of a tree $T$ and
$\textsc{leaves}(T)$ the subset of leaf nodes.

Given a controller (environment)
abstract game tree $T$ a \emph{partial strategy} $Strat: \textsc{nodes}(T) \rightarrow L_u$ 
($Strat: \textsc{nodes}(T) \rightarrow L_c$) labels each node of the tree with the opponent's
action to be played in this node.   
Figure~\ref{fig:strategy} shows an example partial strategy.  The environment 
starts by choosing action $\alpha$.  If the controller plays $a$, the
environment responds with $\beta$ in the next round, and so on. Given a partial
strategy $Strat$, we can map each leaf $l$ of the abstract game tree to
$\langle s',i'\rangle=\textsc{outcome}(\langle s, i\rangle, Strat, l)$ obtained
by playing all controllable and uncontrollable actions on the path from the
root to the leaf.

@@

The algorithm takes a concrete game $G$ with bound $n$ as an implicit argument.
In addition, it takes a player $p$ (controller or environment), state $s$,
bound $k$ and an abstract game tree $T$ and returns a winning partial strategy
for $opponent(p)$, if one exists.  The initial invocation of the algorithm
takes the initial state $\langle I, n\rangle$ and an empty abstract game tree
$\emptyset$.  Initially the solver is playing on behalf of the environment
since that player takes the first move in every step.  The empty game tree does
not constrain opponent moves, hence solving such an abstraction is equivalent
to solving the original concrete game.

The algorithm is organised as a counterexample-guided abstraction refinement
(CEGAR) loop.  The first step of the algorithm uses the \textsc{findCandidate}
function, described below, to come up with a candidate partial strategy for
$T$. If it fails to find a strategy, this means that no winning partial
strategy exists for $T$.  If, on the other hand, a candidate partial strategy
is found, we need to verify if it is indeed winning for $T$.

The \textsc{verify} procedure searches for a \emph{spoiling} counterexample
strategy in each leaf of the candidate partial strategy by calling
\textsc{solveAbstract} for the opponent. The dual solver solves games on behalf
of the controller player.

@@

The way game trees are refined is determined by the formula passed to the SAT
solver in $\textsc{findCandidate}$. This formula contains CNF encodings of all
of the unrolled runs represented by an abstract game tree and the winning
condition of the current player. Runs are encoded by copying the transition
relation for every step in the abstract game. When playing for the controller,
the SAT solver searches for a satisfying assignment to the unfixed label
variables in tree so that none of the runs reaches the error state. The
environment formulation is satisfiable if any run does reach the error state.
The formula is constructed recursively from the root of a tree by
$\textsc{treeFormula}$ (see Algorithm~\ref{alg:treeFormula}).

@@
